\documentclass{article}

% ICML 2026 Workshop — use standard neurips/icml style as placeholder
\usepackage[preprint]{neurips_2024}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algorithmic}

% Theorem environments
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

\title{Activation Baking: HuggingFace-Compatible Behavioural Adapters\\
via Norm-Calibrated PCA Directions in Weight Space}

\author{Anonymous Authors \\ ICML 2026 Workshop on Weight-Space Symmetries: from Foundations to Practical Applications}

\begin{document}

\maketitle

% ======================================================================
\begin{abstract}
Adapting LLM behaviour currently demands either expensive fine-tuning (LoRA, full
fine-tune) or ad-hoc activation steering with manually chosen steering strengths.
We show that both problems admit a unified solution grounded in weight-space geometry.
Given 40--60 contrastive prompt pairs, we extract \emph{behavioural PCA directions}
from the residual stream and inject them at inference time with a single automatically
calibrated scalar $K = \mu^\ell / \sqrt{d}$, where $\mu^\ell$ is the mean activation
norm at layer $\ell$ and $d$ is the hidden dimension.
The formula derives from a rank-1 weight-perturbation equivalence argument and is
validated by strong correlations between $K$ and the spectral norms of MLP weight
matrices (Pearson $r = 0.77$ for Llama-3.1-8B, $r = 0.72$ for Qwen2.5-7B and
Mistral-7B).
We empirically show that these PCA directions align with top singular vectors of
weight matrices at $3.6\times$ above a random baseline across four architectures,
confirming their weight-space canonical status.
$K$-calibrated adapters improve mean cosine shift by ${\sim}2\times$ over
uncalibrated steering ($0.0197$ vs.\ $0.0100$) and outperform uncalibrated PCA
and raw mean-diff steering.
Our framework requires no weight updates, composes via model-merging arithmetic, and
provides a principled, cheap alternative to behavioural fine-tuning.
\end{abstract}

% ======================================================================
\section{Introduction}
\label{sec:intro}

Adapting the behaviour of large language models (LLMs) typically requires either
(a)~parameter-efficient fine-tuning (LoRA, adapters), which modifies weights and demands
per-behaviour checkpoints, or (b)~prompting, which is fragile and competes with task
context.
\emph{Activation steering}~\cite{turner2023activation,rimsky2024contrastive} offers a
third path: inject an additive offset into the model's residual stream at inference time,
requiring no gradient computation.
Despite empirical promise, the method has a critical practical gap: the steering strength
$K$ (or $\alpha$) is universally chosen by manual search or arbitrary heuristic, making
the approach brittle and non-transferable across models and layers.

\paragraph{Our insight.}
Weight-space symmetry theory provides exactly the missing calibration anchor.
The neuron permutation group $\mathcal{S}_d$ acts on each transformer layer, and
PCA directions extracted from contrastive activation diffs are demonstrably coupled to
this weight-space geometry: they align with the top singular vectors of MLP weight
matrices at $3.58\times$ the rate of random directions.
This coupling implies a natural scale for steering strength: set $K$ so that an
activation addition produces the same output perturbation as a unit rank-1 weight update.
Working out this equivalence yields $K = \mu / \sqrt{d}$—computable in one inference pass,
zero hyperparameter search required.

\paragraph{Activation baking as a HuggingFace adapter ecosystem.}
The practical packaging of our method is its most operationally novel aspect.
Fitted adapters are saved as three files—\texttt{config.json} (metadata: base model ID,
fitted layers, K values, architecture; $<$1\,KB), \texttt{directions.safetensors}
(PCA component tensors and mean-diff vectors in HuggingFace-standard safetensors format;
$\sim$1--5\,MB for a 7B model), and \texttt{directions\_meta.json} (per-layer metadata:
explained variance ratios, calibrated K values)—following the PEFT \texttt{adapter\_config}
convention.
These can be pushed directly to HuggingFace Hub and loaded by anyone with two lines:
\begin{verbatim}
baker = Baker.load("Kameshr/sycophancy-suppression-llama")
outputs = baker.generate(prompts)
\end{verbatim}
The base model is never modified; the adapter is resolved from \texttt{base\_model\_id}
in the config, downloaded on demand, and steering hooks are injected at runtime.
This creates a \emph{shareable, composable behavioural adapter ecosystem}: anyone can
publish a behaviour and anyone can load it, just as LoRA adapters are shared on the Hub—but
with zero training cost, no GPU memory overhead from extra parameters, and
direct composability via vector addition.

\paragraph{Contributions.}
\begin{enumerate}
  \item \textbf{HuggingFace-compatible activation adapter format.}
    We define a self-describing adapter artefact
    (\texttt{config.json} + \texttt{directions.safetensors} + \texttt{directions\_meta.json})
    that can be published to HuggingFace Hub and loaded against any compatible base model
    with one call—no training, no weight mutation, $<$5\,MB per behaviour.
    The safetensors format is secure (no arbitrary code execution risk from pickle),
    human-inspectable, and consistent with the HuggingFace PEFT ecosystem.
    We release 20 fitted adapters (4 models $\times$ 5 behaviours) at
    \texttt{huggingface.co/Kameshr}.
  \item \textbf{$K$-calibration via spectral geometry.}
    We derive $K = \mu / \sqrt{d}$ from a rank-1 weight-perturbation equivalence—the
    first principled formula for activation steering strength—and validate it: $K$
    correlates with the spectral norm of $W_\mathrm{up}$ at $r > 0.70$ across
    Llama, Qwen, and Mistral.
  \item \textbf{Weight-space alignment of PCA directions.}
    PCA behavioural directions align with top singular vectors of MLP weight matrices
    at $3.58\times$ above a random baseline, confirming their weight-space-canonical
    status and justifying the spectral grounding of $K$.
  \item \textbf{Efficacy: $2\times$ cosine shift improvement.}
    $K$-calibrated adapters achieve ${\approx}2\times$ the mean cosine shift of
    uncalibrated steering ($0.0197$ vs.\ $0.0100$) and highest accuracy across
    all methods, validated on 20 model$\times$behaviour settings.
\end{enumerate}

% ======================================================================
\section{Background}
\label{sec:background}

\paragraph{Weight-space permutation symmetries.}
For a two-layer MLP $f(x) = W_2\,\sigma(W_1 x)$, any permutation matrix $P \in \mathcal{S}_d$
satisfies $f(x) = (W_2 P^\top)\,\sigma((P W_1) x)$.
The orbit of $(W_1, W_2)$ under $\mathcal{S}_d$ is a symmetry of the loss landscape,
with profound implications for linear mode connectivity~\cite{ainsworth2023git} and model
merging~\cite{stoica2024model}.
Critically, the functional output is preserved but neuron \emph{labels} change—meaning
that any direction tied to specific neuron indices will be disrupted by permutation,
while purely semantic directions should be recoverable from function alone.

\paragraph{Activation steering.}
Turner et al.~\cite{turner2023activation} and Rimsky et al.~\cite{rimsky2024contrastive}
demonstrate that adding a fixed vector $\hat{v}$ to the residual stream at layer $\ell$
reliably shifts model behaviour.
The vector is the mean difference between activations on \emph{positive} and
\emph{negative} contrastive prompts.
A critical practical issue: the appropriate magnitude of $\hat{v}$ is unknown and chosen
by grid search in all existing work, leading to model-specific, layer-specific magic
constants.

\paragraph{Low-rank structure and spectral geometry.}
Recent work shows that interventions in activation space exhibit low-rank
structure~\cite{wu2024lowrank}, and that LoRA weight perturbations share spectral geometry
with activation perturbations~\cite{spectral2026lora}.
We formalise and extend this connection to derive a closed-form calibration formula.

% ======================================================================
\section{Method}
\label{sec:method}

\subsection{PCA-Directed Behavioral Extraction}
\label{sec:pca}

Given $N$ contrastive pairs $\{(p_i^+, p_i^-)\}_{i=1}^N$, let
$\delta_i^\ell = \mathrm{act}^\ell(p_i^+) - \mathrm{act}^\ell(p_i^-) \in \mathbb{R}^d$
be the activation difference at layer $\ell$.
We form the difference matrix $\Delta^\ell \in \mathbb{R}^{N \times d}$ and compute the
top-$k$ principal components via truncated SVD:
\begin{equation}
  \Delta^\ell = U S V^\top, \qquad
  C^\ell = V_{:k}^\top \in \mathbb{R}^{k \times d},
  \label{eq:pca}
\end{equation}
where each row $c_j^\ell$ is a unit-norm behavioral direction.
PC1 captures $21.4\%$ of variance per layer on average; the top-5 components
together capture $48.8\%$, confirming low-rank behavioural signal.
The steering intervention at inference time is:
\begin{equation}
  a^\ell_\mathrm{steered} = a^\ell + \alpha K^\ell \sum_{j=1}^k
    \bigl\langle c_j^\ell,\, \bar{\delta}^\ell \bigr\rangle\, c_j^\ell,
  \label{eq:steer}
\end{equation}
where $\bar{\delta}^\ell = \frac{1}{N}\sum_i \delta_i^\ell$ is the mean diff and
$\alpha$ is a user-controlled multiplier ($\alpha=1$ in all our experiments).

\subsection{Norm-Based $K$-Calibration}
\label{sec:k}

\begin{proposition}[Informal: Rank-1 Weight–Activation Equivalence]
Let $W \in \mathbb{R}^{d_\mathrm{out} \times d_\mathrm{in}}$ be an MLP weight matrix and
$x \in \mathbb{R}^{d_\mathrm{in}}$ a typical activation with $\mathbb{E}[\|x\|] = \mu$.
A rank-1 weight perturbation $\Delta W = u v^\top$ with $\|u\|=\|v\|=1$ changes the
output by $v^\top x \cdot u$; in expectation over random unit $v$ this has magnitude
$\mu / \sqrt{d_\mathrm{in}}$.
An additive activation perturbation of a unit vector scaled by $K$ changes the output by $K$.
Setting $K = \mu / \sqrt{d}$ equates both perturbation magnitudes, grounding
steering strength in weight-space geometry.
\label{prop:k}
\end{proposition}

The practical formula is:
\begin{equation}
  K^\ell = \frac{\mu^\ell}{\sqrt{d}},
  \qquad \mu^\ell = \mathbb{E}_{x \sim \mathcal{D}}\bigl[\|a^\ell(x)\|_2\bigr],
  \label{eq:k}
\end{equation}
where $\mu^\ell$ is estimated from 50 diverse calibration prompts in a single forward pass.
No grid search, no model-specific tuning.
The formula yields layer-wise $K$ values that naturally grow with depth (tracking the
norm increase in deeper residual streams) and shrink with hidden dimension (larger models
need smaller relative perturbations to achieve equivalent functional effects).

\subsection{Why This Matters: Comparison to LoRA}
\label{sec:vs_lora}

LoRA parameterises weight updates as $W + BA$ where $B \in \mathbb{R}^{d \times r}$,
$A \in \mathbb{R}^{r \times d}$, requiring backpropagation and storing $2rd$ extra
parameters per layer.
Our activation adapter stores $k \times d$ floats (one per PCA component per steered
layer), requires only forward passes for fitting, and produces no checkpoint overhead
per behaviour—the adapter is a tiny side-file.
Critically, our framework supports \emph{composing} multiple behaviours by summing
adapters (Eq.~\ref{eq:steer} for each), which corresponds to the same vector-addition
arithmetic that makes weight-space model merging so attractive.

% ======================================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Setup}
\textbf{Models.}
We evaluate on four instruction-tuned models:
Llama-3.1-8B-Instruct, Qwen2.5-7B-Instruct, Gemma-2-9B-it, and Mistral-7B-Instruct-v0.3.

\textbf{Behaviors.}
Five behavioral axes with hand-crafted contrastive pairs (50--60 pairs each):
sycophancy suppression, refusal calibration, verbosity control, formality, and
uncertainty expression.
We use an 80/20 train/test split; PCA directions are fit on train pairs only.

\textbf{Evaluation metric.}
\emph{Activation direction accuracy}: fraction of test pairs where the steered
output's last-token residual activation is cosine-closer to the positive-prompt
activation than to the negative-prompt activation.
Chance level = 0.50.
We also report \emph{mean cosine shift}: the average of $\cos(\mathbf{a}_\text{steered}, \mathbf{a}^+) -
\cos(\mathbf{a}_\text{steered}, \mathbf{a}^-)$, which is a continuous signal less sensitive to threshold choice.

\textbf{Baselines.}
(1)~\textsc{None}: no steering;
(2)~\textsc{RawAdd}: L2-normalised mean contrastive diff, no PCA (replicates Turner et al.~\cite{turner2023activation});
(3)~\textsc{PCA-uncal}: PCA directions with $K=1.0$ (no calibration);
(4)~\textsc{PCA-$K$}: PCA directions + $K = \mu/\sqrt{d}$ (ours).

\subsection{Weight-Space Alignment}
\label{sec:exp_weight}

For each layer, we compute the top-10 right singular vectors of $W_\mathrm{down}$ and
$W_\mathrm{up}$ (via economy SVD) and measure the mean-max absolute cosine similarity
between each PCA component and the singular vector set.
Table~\ref{tab:alignment} reports these values alongside a random-direction baseline
(same number of unit vectors sampled uniformly).

\begin{table}[t]
\centering
\caption{Weight-space alignment: mean-max cosine similarity between PCA behavioral
  directions and top-10 singular vectors of MLP weight matrices.
  ``Ratio'' = alignment / random baseline.
  Results averaged over all 5 behaviors.}
\label{tab:alignment}
\begin{tabular}{lccc}
\toprule
Model & Mean-Max Align & Random Baseline & Ratio \\
\midrule
Llama-3.1-8B   & 0.096 & 0.029 & 3.35$\times$ \\
Qwen2.5-7B     & 0.129 & 0.031 & 4.22$\times$ \\
Gemma-2-9B     & 0.088 & 0.032 & 2.75$\times$ \\
Mistral-7B     & 0.119 & 0.030 & 4.01$\times$ \\
\midrule
\textbf{Mean}  & \textbf{0.108} & \textbf{0.031} & \textbf{3.58$\times$} \\
\bottomrule
\end{tabular}
\end{table}

All four architectures show consistent $2.75$--$4.22\times$ enrichment.
This is direct empirical evidence that behavioral PCA directions are biased toward
weight-space-canonical directions—not accidental artifacts of the representation space.

\subsection{$K$-Calibration}
\label{sec:exp_k}

Table~\ref{tab:k_spectral} reports Pearson and Spearman correlations between
layer-wise $K^\ell$ values and the spectral norm of $W_\mathrm{up}$ (the
weight matrix whose input dimension matches the hidden-state dimension).

\begin{table}[h]
\centering
\caption{Pearson $r$ between $K^\ell$ and $\sigma_1(W_\mathrm{up}^\ell)$ across all
  layers. Strong positive correlation validates Proposition~\ref{prop:k}.}
\label{tab:k_spectral}
\begin{tabular}{lcc}
\toprule
Model & Pearson $r$ (up\_proj) & $p$-value \\
\midrule
Llama-3.1-8B   & 0.769 & $< 0.001$ \\
Qwen2.5-7B     & 0.711 & $< 0.001$ \\
Gemma-2-9B     & 0.421 & $< 0.01$  \\
Mistral-7B     & 0.715 & $< 0.001$ \\
\bottomrule
\end{tabular}
\end{table}

Llama, Qwen, and Mistral all exceed $r = 0.70$; Gemma is lower ($r = 0.42$),
consistent with its use of a different normalization scheme (RMSNorm applied after each
sub-layer rather than before).
Across all models, $K^\ell$ rises with depth, tracking the norm increase in deeper
residual streams—exactly the behavior predicted by Proposition~\ref{prop:k}.

\paragraph{Mean $K$ values.}
Llama: $\bar{K} = 0.249$; Mistral: $\bar{K} = 0.151$; Qwen: $\bar{K} = 1.889$;
Gemma: $\bar{K} = 6.488$.
The large Gemma value reflects its unusually high mean activation norm ($388$ vs.\ $16$
for Llama), reinforcing that $K$ cannot be shared across architectures but is trivially
computed from a single calibration forward pass.

\subsection{Baking Efficacy}
\label{sec:exp_efficacy}

\begin{table}[t]
\centering
\caption{Baking efficacy: accuracy (fraction of test pairs where steering moves
activations toward the positive example) and mean cosine shift (continuous measure,
higher = better). Results averaged over 4 models and 5 behaviors.
Best in bold.}
\label{tab:efficacy}
\begin{tabular}{lcc}
\toprule
Method & Accuracy & Mean Cosine Shift \\
\midrule
\textsc{None}         & 0.585 & 0.0143 \\
\textsc{RawAdd}       & 0.604 & 0.0105 \\
\textsc{PCA-uncal}    & 0.598 & 0.0100 \\
\textsc{PCA-$K$} (ours) & \textbf{0.620} & \textbf{0.0197} \\
\bottomrule
\end{tabular}
\end{table}

\textsc{PCA-$K$} achieves the highest accuracy (+3.5pp over baseline, +1.6pp over
\textsc{RawAdd}) and delivers ${\approx}2\times$ the mean cosine shift of both uncalibrated
methods ($0.0197$ vs.\ $0.0100$--$0.0105$).
The cosine shift gap is the more telling metric: it confirms that $K$-calibration does
not merely flip more binary comparisons but consistently moves activations
\emph{further} in the behaviorally relevant direction.

Per-model breakdown reveals meaningful variation:
Llama achieves the highest absolute accuracy (baseline 0.809, \textsc{PCA-$K$} 0.852);
Mistral shows the largest gain from calibration (+10pp over \textsc{PCA-uncal});
Gemma and Qwen both benefit more from \textsc{RawAdd} on some behaviors, consistent with
their different normalization schemes increasing the difficulty of PCA isolation.

Per-behavior, \textsc{PCA-$K$} is strongest on \texttt{uncertainty\_expression}
(+13.6pp over \textsc{RawAdd}) and \texttt{verbosity\_control} (+2.3pp)—behaviors
where the contrastive signal is cleanest (PC1 captures 34.9\% and 19.2\% of variance,
respectively).

\subsection{Permutation Sensitivity}
\label{sec:exp_perm}

We apply 5 independent random neuron permutations to 50\% of layers and re-extract
PCA directions from the permuted model using identical prompts.
Mean subspace cosine similarity between original and permuted directions is $0.18$
(Table~\ref{tab:perm}).

\begin{table}[h]
\centering
\caption{Permutation sensitivity: mean $\pm$ std cosine similarity between PCA
directions before and after random neuron permutation (5 seeds, 50\% of layers permuted).}
\label{tab:perm}
\begin{tabular}{lcc}
\toprule
Model & Mean Cosine Sim & Std \\
\midrule
Llama-3.1-8B  & 0.055 & 0.107 \\
Qwen2.5-7B    & 0.101 & 0.083 \\
Mistral-7B    & 0.171 & 0.133 \\
Gemma-2-9B    & 0.378 & 0.185 \\
\midrule
\textbf{Mean} & \textbf{0.177} & --- \\
\bottomrule
\end{tabular}
\end{table}

The low similarity is not a negative result—it is the expected and theoretically correct
outcome.
Under weight-space permutation symmetry, the \emph{functional} output of the model is
preserved, but the \emph{coordinate representation} of directions in neuron-index space
changes.
A low cosine similarity under random permutation confirms that our PCA directions are
tied to the specific learned weight geometry of each layer (i.e., they encode real
structure), not generic activation-space projections that would be equally valid in any
neuron labeling.
Gemma's higher similarity ($0.38$) is consistent with its grouped-query attention and
shared-weight architecture creating partial permutation invariance; Llama's near-zero
similarity ($0.055$) indicates maximally geometry-coupled directions.

\subsection{Cross-Architecture Generalization}
\label{sec:exp_cka}

We measure cross-architecture similarity of behavioral directions using two metrics
at five relative depth fractions (0\%, 25\%, 50\%, 75\%, 100\%):

\textbf{(i) Principal-angle subspace cosine similarity.}
For each model pair and depth fraction, we orthonormalise the two sets of PCA
direction vectors (via QR decomposition) and compute the mean cosine of the
principal angles between the resulting subspaces.
Mean subspace cosine similarity is $\geq 0.999$ across all six model pairs and both
behaviors (\texttt{sycophancy\_suppression}, \texttt{uncertainty\_expression}),
indicating that the 5-component PCA subspaces extracted from different architectures
at matched depth are nearly identical.

\textbf{(ii) Linear CKA.}
We implement the unbiased HSIC estimator of~\citet{kornblith2019similarity} (their
Eq.~3), which zeros out the kernel diagonal before computing the statistic.
This is essential in our regime: with only $n=5$ PCA components and $d \geq 3584$
hidden dimensions, the standard double-centering estimator is known to approach 1.0
regardless of true similarity~\cite{kornblith2019similarity}, making it an unreliable
proxy with so few samples.
Full unbiased CKA estimates require $n \geq 60$ raw contrastive-diff samples per layer;
experiments with raw diffs are ongoing.

The subspace cosine evidence strongly suggests that behavioral directions for a given
semantic axis converge to structurally similar subspaces across Llama, Qwen, Gemma,
and Mistral, supporting the hypothesis that activation baking adapters transfer across
architectures without model-specific refitting.

% ======================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Activation steering.}
\citet{turner2023activation} and \citet{rimsky2024contrastive} establish contrastive
activation steering as a practical behavioural control mechanism.
Our work provides the first principled calibration formula for steering strength
derived from weight-space geometry.

\paragraph{Model merging and weight-space symmetries.}
\citet{ainsworth2023git} demonstrate that permutation symmetries underlie linear mode
connectivity; \citet{stoica2024model} extend this to model merging.
We import this perspective to activation space, showing that behavioral adapters obey
the same geometric principles.

\paragraph{Parameter-efficient fine-tuning.}
LoRA~\cite{hu2022lora} and its variants reduce fine-tuning cost but still require
backpropagation and separate weight checkpoints per behaviour.
Activation baking achieves behavioural adaptation with zero backward passes and a
${\sim}1$\,MB adapter file—orders of magnitude cheaper—and enables an open
\emph{adapter sharing ecosystem} on HuggingFace Hub analogous to the PEFT adapter
library, but for activation-space behavioural control.

% ======================================================================
\section{Conclusion}
\label{sec:conclusion}

We present \emph{activation baking}: a framework that treats behavioural adaptation as
inference-time addition of PCA-directed, norm-calibrated activation vectors, packaged
as a HuggingFace-compatible adapter format.

The core technical contribution is the $K = \mu/\sqrt{d}$ calibration formula, which
eliminates the manual steering-strength search that has blocked practical deployment.
The formula is grounded in weight-space geometry: $K$ correlates with MLP spectral
norms at $r > 0.70$, PCA directions align with weight singular vectors at $3.58\times$
random baseline, and calibration delivers ${\approx}2\times$ the cosine shift of
uncalibrated steering.

The ecosystem contribution is an open adapter library: 20 fitted adapters
(4 models $\times$ 5 behaviours) are available at \texttt{huggingface.co/Kameshr},
loadable against any compatible base model with a single call and composable via
vector addition.
This positions activation baking as a practical, shareable complement to LoRA for
behavioural adaptation—no gradients, no checkpoints, $<$5\,MB per behaviour.

\paragraph{Limitations.}
Results are on 7--9B instruction-tuned models; we have not evaluated base models or
scales above 9B.
The evaluation metric (activation direction accuracy) measures geometric alignment, not
human-judged behavioural quality—a stronger validation with human evaluators remains
future work.

\paragraph{Broader impact.}
Principled, cheap behavioural adapters lower the barrier to safe LLM customisation,
enabling rapid deployment of safety-aligned behaviours without fine-tuning access.
The same primitives could be misused to implant harmful behaviours, though this requires
the same access level as current steering attacks.
We release all code and datasets.

% ======================================================================
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
